---
title: "R Notebook"
output: html_notebook
author: Kaz Shakir
---

```{r}
# setup environment
# 
library(magrittr)
library(rpart)
library(dplyr)
library(ggplot2)
library(scales)
library(gridExtra)

# load and clean the initial data
# 
# load the csv file into a data.frame
# convert the Date,Started field into a Date type
# and make the County names all caps
# 
icData <- read.csv(file = '20181009.csv')
icData$Date.Started <- as.Date(as.character(icData$Date.Started), format = "%m/%d/%Y")
icData$County <- toupper(icData$County)

# create a variable for the estimatedAge, and another
# for the jobTerm (the number of years person has been in the job)
# 
icData$estimatedAge <- as.integer(format(Sys.Date(), "%Y")) - icData$Year.of.Birth
icData$jobTerm <- as.integer(format(Sys.Date(), "%Y")) - as.integer(format(icData$Date.Started, "%Y"))

# create a subset of the data that only includes
# elected officials, and full-time employees
# 
modelSubset <- 
  subset(icData, 
         ((icData$Employment.Status == 'Elected' ||
            icData$Employment.Status == 'Full Time')) &
           (icData$Annual.Salary >= 12000)) %>%
  droplevels()

# calculate some descriptive statistics by Job Category
# 
a1 <- modelSubset %>%
  group_by(EEO.Job.Cat.Desc) %>%
  summarise(low = min(Annual.Salary), high = max(Annual.Salary), avg = mean(Annual.Salary),
            stdDev = sd(Annual.Salary), cv = sd(Annual.Salary) / mean(Annual.Salary)) %>%
  arrange(avg)


```

### Salary statistics by Job Description category:
```{r}
print(a1)
```

* Observations: 
  + The Elected Officials seem to have a very wide dispersion of salaries, as demonstrated by the cv ratio (whcih is the standard deviation divided by the average).
  + The Officials and Administrators categry also seems to have a wide dispersion of values, ranging from a low of $15,000 to a high of $319,200.
  + The "Officials and Administrators", "Professionals", and "Elected Official" categories all seem to have particularly high paid positions
  
### A box and whisker chart of the distribution of salaries by job description category, excluding 
### the Elected Officials, Professionals, and Officials and Administrators
```{r}
ggplot(data = modelSubset[(modelSubset$EEO.Job.Cat.Desc != 'Elected Officials') &
                            (modelSubset$EEO.Job.Cat.Desc != 'Officials and Administrators') &
                            (modelSubset$EEO.Job.Cat.Desc != 'Professionals'),],
       aes(x = EEO.Job.Cat.Desc, y = Annual.Salary, ymax = 100000, ymin = 0)) +
  geom_boxplot() +
  geom_violin(alpha = 0.4) +
  scale_y_continuous(labels = comma) +
  theme(axis.text.x = element_text(size = 10, angle = 25),
        panel.background = element_rect(fill = FALSE),
        panel.grid.major = element_line(color = 'gray'))
  
```
* Observations:
  + Based on the levels of the various boxes, it appears that there are two prevalent groups
    + First, the Support group that includes: Administrative Support, Para-Professionals, and Service Maintenance
    + Second, the Skilled Labor group that includes: Protective Service Workers, Skilled Craft Workers, and Technicians
    
### A box and whisker chart for the Elected Officials, Officials and Administrators, and Professionals
```{r}
ggplot(data = modelSubset[(modelSubset$EEO.Job.Cat.Desc == 'Elected Officials') |
                            (modelSubset$EEO.Job.Cat.Desc == 'Officials and Administrators') |
                            (modelSubset$EEO.Job.Cat.Desc == 'Professionals'),],
       aes(x = EEO.Job.Cat.Desc, y = Annual.Salary, ymax = 325000, ymin = 0)) +
  geom_boxplot() +
  geom_violin(alpha = 0.4) +
  scale_y_continuous(labels = comma) +
  theme(axis.text.x = element_text(size = 10, angle = 25),
        panel.background = element_rect(fill = FALSE),
        panel.grid.major = element_line(color = 'gray'))
  

```
The Officials and Administrators and Professionals categories could be grouped into a single category
called Management - they seem to be similar enough.

Elected officials, however, seem to have a very odd profile: a wide range, but the median is all the way to the bottom.

### An ordered scatter plot of the Elcted Officials category
```{r}
plot(sort(modelSubset$Annual.Salary[modelSubset$EEO.Job.Cat.Desc == 'Elected Officials']),
     ylab = "Annual Salary", xlab = "Elected Official")
```
Now, it seems clear that the elected officials seem to group into three distinct categories,
with most of them below \$25k and then a small group around \$130 or \$140, and 
a thrid group at \$170 or \$180.

So, it seems that there are three distinct groups:
  + Support
  + Skilled Labor
  + Management 
  
```{r}
# add a column to the dataframe to represent the three major groups and Elected Officials
# as the fourth group.
# 
modelSubset$group <-
  as.factor(
    apply(modelSubset, MARGIN = 1, FUN = function(x) {
      if (x["EEO.Job.Cat.Desc"] %in% 
          c('Administrative Support', 'Para-Professional', 'Service Maintenance')) 
        {return ('Support')}
      if (x["EEO.Job.Cat.Desc"] %in% 
          c('Protective Service Workers', 'Skilled Craft Workers', 'Technicians')) 
        {return ('Skilled Labor')}
      if (x["EEO.Job.Cat.Desc"] %in% 
          c('Officials and Administrators', 'Professionals')) 
        {return ('Management')}
      if (x["EEO.Job.Cat.Desc"] == 'Elected Officials') 
        {return ('Elected')}
    })
  )
```

### Draw denisty of the distribution for each of the four categories
```{r}
ggplot(modelSubset, 
       aes(Annual.Salary, group = group, fill = group, alpha = 0.25)) +
  geom_density()+
  scale_x_continuous(label = comma) +
  ggtitle("All Groups") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(modelSubset[modelSubset$group == 'Support',], 
       aes(Annual.Salary, fill = 4, alpha = 0.25)) +
  geom_density() +
  scale_x_continuous(label = comma) +
  ggtitle("Salary Disribution For Support") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(modelSubset[modelSubset$group == 'Skilled Labor',], 
       aes(Annual.Salary, fill = 4, alpha = 0.25)) +
  geom_density() +
  scale_x_continuous(label = comma) +
  ggtitle("Salary Disribution For Skilled Labor") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(modelSubset[modelSubset$group == 'Management',], 
       aes(Annual.Salary, fill = 4, alpha = 0.25)) +
  geom_density() +
  scale_x_continuous(label = comma) +
  ggtitle("Salary Disribution For Management") +
  theme_minimal() +
  theme(legend.position = "none")

```
Each of the groups seems to be normally distributed with some outliers.  The Support group seems
to have a small set of jobs in a low paid category; and the other groups seem to have a few on
the very high end of the scale.

```{r}
# build a function to create summary statistics for evaluation
# 
summaryStats <- function(df, grp) {
  # calculate some summary statistics for the grp group by Ethnic code
  # first, determine the number employed in each ethnic group
  # second, determine the average salary by ethnic group
  # 
  result <-
    df[df$group == grp,] %>% 
    group_by(Ethnic.Code.Description) %>% 
    summarise(num_of_people = n(), avg_salary = mean(Annual.Salary)) %>% 
    arrange(num_of_people)
  
  # determine the overall average salary for the grp group
  # and then calculate the amount each ethnic group is either 
  # above or below that average
  #
  AvgSal <- mean(df$Annual.Salary[df$group == grp])
  result$avg_salary_relative_to_average <-
    result$avg_salary - AvgSal
  
  return(result)
}

# build a function to create two bar charts for the summarize results
# that were produced by the summaryStats function
# 
producePlots <- function(smry, grp, lmts1, brks1, lmts2, brks2) {
  # plot a bar chart of the number of people employed in each ethnic category
  # in the grp group, and a second graph showing the average salary
  # by each ethnic group relative to the average for the grp group
  # 
  p1 <-
    ggplot(smry, 
         aes(reorder(Ethnic.Code.Description, num_of_people), num_of_people)) +
    scale_y_continuous(limits = lmts1, 
                       breaks = brks1,
                       labels = comma) +
    geom_col() + labs(x = NULL) + coord_flip() +
    ggtitle(paste0("Number Of Individuals In the ", grp, 
                   " Group \nBy Ethnic Description")) +
    theme_minimal()
  
  p2 <- 
    ggplot(smry, 
         aes(reorder(Ethnic.Code.Description, num_of_people), avg_salary_relative_to_average,
             fill = as.factor(sign(avg_salary_relative_to_average)))) +
    scale_y_continuous(limits = lmts2, 
                       breaks = brks2,
                       labels = comma) +
    geom_col() + labs(x = NULL) + coord_flip() +
    ggtitle(paste0("In the ", grp, " group, how much does the annual salary \n",
                   "deviate from the average for each Ethnic Description")) +
    theme_minimal() +
    theme(legend.position = "none")
  
  return(list(p1 = p1, p2 = p2))
  
}


supportSummary <- summaryStats(modelSubset, 'Support')
supportPlots <- producePlots(supportSummary, 'Support', 
                             lmts1 = c(0,1400), brks1 = seq(0, 1400, 400),
                             lmts2 = c(-5000,5000), brks2 = seq(-5000,5000,2000))

grid.arrange(supportPlots$p1, supportPlots$p2, ncol = 1)

```

Do the same for the skilled labor group

```{r}
skilledSummary <- summaryStats(modelSubset, 'Skilled Labor')
skilledPlots <- producePlots(skilledSummary, 'Skilled Labor', 
                             lmts1 = c(0,2600), brks1 = seq(0, 2600, 600),
                             lmts2 = c(-7000,7000), brks2 = seq(-7000,7000,2000))

grid.arrange(skilledPlots$p1, skilledPlots$p2, ncol = 1)

```

Do the same calcs for the Management group

```{r}
managementSummary <- summaryStats(modelSubset, 'Management')
managementPlots <- producePlots(managementSummary, 'Management', 
                             lmts1 = c(0,2200), brks1 = seq(0, 2200, 400),
                             lmts2 = c(-8000,8000), brks2 = seq(-8000,8000,2000))

grid.arrange(managementPlots$p1, managementPlots$p2, ncol = 1)

```

